---
title: "Prediction of Exercise Manner"
author: "Duo Sun"
date: "12 August 2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, collapse = FALSE)
```

## 1 Introduction

This goal of this study is to predict the manner (noted as A, B, C, D, E in the dataset) in which 6 participants did the exercise. 

The features used as predictors for this purpose include x-,y-,z-accelerometers on the belt, forearm, arm, and dumbell. Three models including LDA,  CART, and kNN are built, and repeated cross-validation method is used to evaluate the performance and select the optimal parameters for each model. 

Among the built models, kNN gave the best performanc with an average prediction accuracy of 89.6%. The performance of the kNN model is further evaluated on validation data, which gave a prediction accuracy of 91.5%.

## 2 Load data

The dataset "pml-training.csv" containing 19622 observations with 160 variables are used for training and validating models. Among these 159 variable, 12 features that are believed to be directly related with exercise manner including x-,y-,z-accelerometers on the belt, forearm, arm, and dumbell of 6 participants are selected to predict the manner of the exercise.
```{r, results='hide', echo=FALSE}
load("PML_final.RData")
```

```{r}
# Load libraries
library(caret)
library(ggplot2)
library(lattice)
library(corrplot)

# Load dataset
pmldata <- read.csv("pml-training.csv",header = TRUE)

# Remove variables containing NA
pmldata <- pmldata[,colSums(is.na(pmldata))==0]

# Extract features including accelerometers x,y,z on the belt, forearm, arm, and dumbell of 6 participants.  
var <- c("accel_belt_x","accel_belt_y","accel_belt_z",
         "accel_arm_x","accel_arm_y","accel_arm_z",
         "accel_dumbbell_x","accel_dumbbell_y","accel_dumbbell_z",
         "accel_forearm_x","accel_forearm_y","accel_forearm_z",
         "classe")
dataset <- pmldata[,var]
```

## 3 Split the data into training/testing dataset and validation dataset

The dataset are split into two part: 80% dataset is used for training models and estimating their prediction accuracy; 20% dataset is used to evaluate the built model. The distributions of different manners in training dataset and validation dataset are shown below and are very well matched.

```{r}
set.seed(0)
inTrainTest  <- createDataPartition(y=dataset$classe,p=0.80,list=FALSE)
validation   <- dataset[-inTrainTest,]
dataset      <- dataset[inTrainTest,]
dim(dataset)
cbind(freq=table(dataset$classe), percentage=prop.table(table(dataset$classe)) * 100)
dim(validation)
cbind(freq=table(validation$classe), percentage=prop.table(table(validation$classe)) * 100)
```

## 4 Data summary and visualization

The follow figures show range of each selected variable under different manner, and the correlation between variables, respectively.

```{r, fig.width = 7, fig.height = 7}
# Summarize data
dim(dataset)
str(dataset)
# Data visualization
featurePlot(x=dataset[,1:12], y=dataset[,13], plot="box")
# Calculate correlations between predictors
correlations <- cor(dataset[,1:12])
# Create correlation plot
corrplot(correlations, method="circle")
```

## 5 Model building

Three different models including LDA,  CART, and kNN are built. Repeated cross-validation is used to evaluate the performance of models and select the optimal parameters for each model. 

```{r, eval= FALSE, fig.width = 7, fig.height=7}
# Run algorithms using 10-fold cross validation, repeated 3 times
fitControl <- trainControl(method = "repeatedcv",number = 10,repeats = 3)
metric <- "Accuracy"
tuneLength <- 20
# LDA
set.seed(0);fit.lda  <- train(classe~., data=dataset, preProcess = c("center", "scale"), method="lda", metric=metric, trControl=fitControl, tuneLength = tuneLength)
# CART
set.seed(0);fit.cart <- train(classe~., data=dataset, preProcess = c("center", "scale"), method="rpart", metric=metric, trControl=fitControl, tuneLength = tuneLength)
# kNN
set.seed(0);fit.knn  <- train(classe~., data=dataset, preProcess = c("center", "scale"), method="knn", metric=metric, trControl=fitControl, tuneLength = tuneLength)
```

The cross-validation results show that kNN gives the best performace, an average accuracy of 89.6%.

```{r, fig.width = 7, fig.height=7}
# Compare LDA, CART, and kNN
results <- resamples(list(lda=fit.lda, cart=fit.cart, knn=fit.knn))
summary(results)
dotplot(results)
# Summarize Best Model
print(fit.knn)
plot(fit.knn)
```

## 6 Model evaluation

The built kNN model is further evaluated on validation dataset, which shows a predication accuracy of 91.5%, close to the estimated result shown in section 5.

```{r}
# Validate model accuracy using testing data
validation_pred <- predict(fit.knn, newdata = validation)
confusionMatrix(validation_pred, validation$classe)
mean(validation_pred == validation$classe)
```

## 7 Predict new data with the model

The built kNN model is used to predict the new dataset "pml-testing.csv". The prediction result is given below.

```{r}
newdata <- read.csv("pml-testing.csv",header = TRUE)
newdata <- newdata[,var[-13]]
newdata_pred <- predict(fit.knn, newdata = newdata)
newdata_pred
```
